Functional Pearl: the Proof Search Monad

### 2015-10-13 paper/main.tex
As submitted.

@@

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    morekeywords={above,abstract,adopts,alias,and,as,assert,begin,below,builtin,consumes,data,do,downto,duplicable,dynamic,else,empty,end,exclusive,explain,fact,fail,flex,for,from,fun,give,if,in,let,match,mutable,open,pack,perm,preserving,rec,take,taking,then,to,type,unknown,val,value,while,with,witness},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

In order to show the validity of this judgement, one usually builds a proof
derivation using rules from the logic. In our case, the rules are given in
\fref{logic}; $[x/y]P$ means ``substitute $x$ with $y$ in $P$''. For
instance, proving Equation \ref{ex1} requires applying \Rule{ExistsE}, then
\Rule{Refl}.

@@

The library has been used, in a preliminary form, to implement the core of the
Mezzo type-checker. This paper presents a cleaned-up, isolated version of this
library that exposes a proper interface using monads and domain-specific
combinators.

@@

The logic we present is a much simplified version of the logic (type system) of
Mezzo. In this document, we only mention the right-exists quantifier. General
systems such as Mezzo have all four possible combinations of left/right
exists/forall.
%
The right-elimination of existential quantifiers, or the left-elimination of
universal quantifiers gives \emph{flexible variables}, while the
right-elimination of universal quantifiers, or the left-elimination of
existential quantifiers gives universally-quantified variables, also called
\emph{rigid variables}.

In order to simplify the problem, we assume that all existential variables have
been introduced as flexible variables already. That way, we won't be
sidetracked, talking about binders and the respective merits of De Bruijn
\emph{vs.} locally nameless. Furthermore, we assume that all instantiations of
flexible variables are legal. This is not true in general: for instance, if the
goal is $\forall x, \exists \f y, \forall z.\ P$, picking $\f y = z$ makes no
sense. Mezzo forbids this choice using \emph{levels}; in the present document,
we skip this discussion altogether and assume that ``all is well''. Finally,
although in a general setting, several rules may trigger for a given goal (this
is the case in Mezzo), the algorithmic set of rules we use is syntax-driven: the
syntactic shape of the goal determines which rule should be applied.

@@

We implement proof search in OCaml (\fref{formulas}); we implement substitutions
using a union-find data structure. The data type of formulas
is self-explanatory. Variables are implemented as equivalence classes in the
\emph{persistent} union-find data structure, which the module \li+P+ implements.
The $V, \sigma$ parameters in our rules are embodied by the \li+state+ type; just
like the $\sigma$ parameter is chained from one premise to another (\Rule{And}),
\li+state+ is an input and an output to the solver. Just like the $\sigma$ parameter
in the rules, a \li+state+ of the persistent union-find represents
a set of equations between variables. The algorithmic rules mentioned a
theoretical $\sigma$ parameter; the \li+state+ is our specific implementation
choice.

@@

\fref{solver} implements a solver for our minimal problem; written within the
\li+MOption+ monad, it returns either \li+Some state+ (in case a successful
substitution has been found), or \li+None+ if no solution exists. The solver is
complete.

@@

We previously used the \li+>>=+ operator from the \li+MOption+ monad in order to
chain premises (\fref{solver}). We now need a new operator, that not only
\emph{binds} the result (i.e. stops evaluating premises after a failure, as
before), but also \emph{records} the premises in sequence, in order to build a
proper derivation. The former is still faithfully implemented by the option
monad; the latter is implemented by the writer monad.

@@

ZOMG TODO

@@

\bibliographystyle{abbrvnat}
\bibliography{local}

>>>>

\usepackage{fontspec}
  \defaultfontfeatures{Mapping=tex-text}
\usepackage{ucharclasses}
  \setTransitionsFor{MathematicalOperators}{\begingroup\fontspec{DejaVu Sans}[Scale=MatchLowercase]}{\endgroup}
\usepackage{xunicode}

\usepackage{xltxtra}

@@

\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{listings}
  \def\li{\lstinline}
  \lstset{
    language=caml,
    flexiblecolumns=false,
    showstringspaces=false,
    basicstyle=\ttfamily,
    framesep=5pt,
    numberstyle=\tiny,
    numbersep=5pt,
    escapeinside={(*}{*)},
    deletekeywords={value},
  }
\usepackage{minted}
  \newminted[ocaml]{ocaml}{mathescape,fontsize=\small}
\usepackage{mathpartir}
  % Improvements to Didier's mathpartir package: make all rule names and
  % references use textsc, and also make references to rules done using the \Rule
  % command clickable in a PDF.
  \let\TirName\textsc
  \renewcommand{\RefTirName}[1]{\hypertarget{#1}{\TirName {#1}}}
  \renewcommand{\DefTirName}[1]{\hyperlink{#1}{\TirName {#1}}}
  \let\DefRule\RefTirName
  \let\Rule\DefTirName
  % \let\Rule\textsc

@@

In order to show the validity of this judgement, one usually builds a proof
derivation using rules from the logic. In our case, the rules are given in
\fref{logic}, where $[x/y]P$ means ``substitute $x$ with $y$ in $P$''. For
instance, proving Equation \ref{ex1} requires applying \Rule{ExistsE}, then
\Rule{Refl}.

@@

The library has been used, in a preliminary form, to implement the core of the
Mezzo type-checker~\cite{protzenko-phd-14}. This paper presents a cleaned-up,
isolated version of this library that exposes a proper interface using monads
and domain-specific combinators.

@@

The logic we present is a much simplified version of the logic (type system) of
Mezzo~\cite{pottier-protzenko-13}. In the present document, we only mention the
right-exists quantifier. General systems such as Mezzo have all four possible
combinations of left/right exists/forall.
%
The right-elimination of existential quantifiers, or the left-elimination of
universal quantifiers gives \emph{flexible variables}, while the
right-elimination of universal quantifiers, or the left-elimination of
existential quantifiers gives universally-quantified variables, also called
\emph{rigid variables}.

In order to simplify the problem, we assume that all existential variables have
been introduced as flexible variables already. That way, we won't be
sidetracked, talking about binders and the respective merits of De Bruijn
\emph{vs.} locally nameless. Furthermore, we assume that all instantiations of
flexible variables are legal. This is not true in general: for instance, if the
goal is $\forall x, \exists \f y, \forall z.\ P$, picking $\f y = z$ makes no
sense. Mezzo forbids this choice using \emph{levels}~\cite{pottier-remy-emlti}; in the present document,
we skip this discussion altogether and assume that ``all is well''. Finally,
although in a general setting, several rules may trigger for a given goal (this
is the case in Mezzo), the algorithmic set of rules we use is syntax-driven: the
syntactic shape of the goal determines which rule should be applied.

@@

We implement proof search in OCaml~\cite{ocaml} (\fref{formulas}); we implement substitutions
using a union-find data structure~\cite{cormen-en,tarjan-75}. The data type of formulas
is self-explanatory. Variables are implemented as equivalence classes in the
\emph{persistent} union-find data structure, which the module \li+P+ implements.
The $V, \sigma$ parameters in our rules are embodied by the \li+state+ type; just
like the $\sigma$ parameter is chained from one premise to another (\Rule{And}),
\li+state+ is an input and an output to the solver. Just like the $\sigma$ parameter
in the rules, a \li+state+ of the persistent union-find represents
a set of equations between variables. The algorithmic rules mentioned a
theoretical $\sigma$ parameter; the \li+state+ is our specific implementation
choice.

@@

\fref{solver} implements a solver for our minimal problem; since we perform
computations that either return a result of a failure, the code leads itself
well to an implementation using
monads~\cite{wadler-comprehending-92,wadler-essence-92}, in our case, the
\li+MOption+ monad. The \li+Some state+ is for success, meaning a substitution
has been found, while the \li+None+ case means no solution exists. The solver is
complete.

@@

We previously used the \li+>>=+ operator from the \li+MOption+ monad in order to
chain premises (\fref{solver}). We now need a new operator, that not only
\emph{binds} the result (i.e. stops evaluating premises after a failure, as
before), but also \emph{records} the premises in sequence, in order to build a
proper derivation. The former is still faithfully implemented by the option
monad; the latter is implemented by the writer monad~\cite{jones-95-}.

@@

An article titled ``The Proof Monad'' already exists~\cite{kirchner-munoz-10};
in spite of the closely related title, the article is concerned with a slightly
different problem, namely giving an operational semantics to tactic languages
used in theorem provers. In that sense, the article is related to
Mtac~\cite{ziliani-13}, which is also concerned with a proper monad for writing
tactics in Coq.

Hedges~\cite{hedges-14} compares various explorations monads, notably using the
continuation monad, the selection monad~\cite{escardo-2010} and their respective
monad transformers. The main focus of the article seems to be the relationship
between backtracking and game theory.

Hinze~\cite{hinze-00} shows how to use the backtracking monad transformer, i.e.
add backtracking to any existing monad. It would be interesting to determine
whether our library can be re-implemented using a backtracking monad
transformer, rather than the writer transformer applied to the monad of
non-determinism. The (draft) version of the library used in Mezzo also builds
failed derivations (as error messages) that list all attempted proofs, along
with the first premise that failed; doing so would not be possible using
exceptions.

The \li+choice+ operator is related to polarization and
focusing~\cite{liang-miller-07}. For instance, in the problem $\f y = x \vee \f
y = z$, depending on which side of the disjunction the algorithm considers first,
the outcome is going to be different. This is analogous to a synchronous phase
(where the order of the rules matters, and where a particular choice may have
consequences on the rest of the search). Similarly, one may swap premises
chained by the \li+>>=+ operator, as the order doesn't matter. This is analogous
to a asynchronous phase.

@@

\bibliographystyle{plain}
\bibliography{english,local}
